### 2022-09-09 HW1/HW1.Rmd
start hw1

---
title: "HW 1"
author: "Hayden Atchley"
date: "9 September 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```

# 1
The systems boundary for SLC Int'l Airport is below:

![SLC Int'l Airport systems boundary.](image/boundary.png)

I think most of the elements of the diagram are fairly straightforward, where emissions sources produced directly by the airport are entirely inside the boundary, though a few elements are worth mentioning specifically. I considered personal travel to be a part of the trip in which a dropoff/pickup occurs, so the idling and minimal driving that happens with a dropoff/pickup is entirely within the airport boundary, while the transportation to and from the airport is not. Additionally, while the airport _does_ in a sense generate flights, the in-flight emissions themselves are outside the boundary of the airport (the emissions during landing, takeoff, and taxiing are entirely in the boundary).

Another, perhaps easier, way to think about it (and how I thought about it) is a question of control. What emissions sources does the airport have total control over, and what sources are only partially under their control? The airport can, for example, control how much electricity it uses, but not the electricity to emissions ratio of the source. SLC International can (in theory at least) control the emissions of a plane while it's on the ground, but once the plane is in the air the airport has no control over the emissions.

# 2
The causal loop diagram is below:

![Causal loop diagram exploring gas and electric vehicles.](image/loop.png)

This analysis is far from perfect, but it gives a good idea of the relationships at play. I made a few assumptions:

 - An increase in gas price would lower the demand for (and therefore number of) gas-powered vehicles. This is probably not true in the short term, but is probably true in the long term.
 - Emissions regulations are likely to have an impact on fuel economy, though I'm not sure in which direction. Certain regulations may cause manufacturers to use engines that burn more cleanly, but that might have a negative impact on fuel economy. However, there may also be improvements in engine emissions that _increase_ fuel economy as well.
 - I am assuming that as the supply of rare earth metals decreases, battery technology will improve in a way that uses less of them. This may or not be the case, but I think it's likely.

# 3

### 2022-09-13 HW1/HW1.Rmd
update HW1

@@
 )
+
+library(tidyverse)
+library(gridExtra)
 ```
@@
 # 3
+I used the logistic growth equation to model electric vehicle shares. The equation is given as
+\[f(t) = \frac{e^{a+bt}}{1+e^{a+bt}}\]
+where $f(t)$ is the fraction of electric vehicles, $t$ is the year, and $a$ and $b$ are coefficients.
+
+Data from 2011\textendash 2019 were obtained from [https://afdc.energy.gov/data/] and [https://www.epa.gov/automotive-trends/explore-automotive-trends-data#SummaryData].
+The adfc data include total EV sales by year, and the epa data include information on all car production.
+I am only considering fully-electric vehicles as EVs for this analysis (not plug-in hybrid EVs).
+
+```{r}
+cars <- read_csv("HW1/data/table_export.csv") %>% 
+	mutate(`Model Year` = as.integer(`Model Year`),
+				 `Production (000)` = as.integer(`Production (000)`)) %>% 
+	filter(`Regulatory Class` == "Car",
+				 `Vehicle Type` == "All Car",
+				 Manufacturer == "All",
+				 `Model Year` %in% 2011:2019) %>% 
+	mutate(`Model Year` = as.character(`Model Year`))
+
+evs <- read_csv("HW1/data/10567_pev_sales_2-28-20.csv") %>% 
+	filter(Type == "EV") %>% 
+	select(-Vehicle, -Type) %>% 
+	colSums() %>% 
+	{tibble(Year = names(.),
+					EVs = .)} 
+```
+
+Plotting the sale of EVs and total vehicles gives:
+
+```{r}
+ev_plot <- evs %>% 
+	ggplot(aes(x = Year, y = EVs/1000)) +
+	geom_point() +
+	ylab("EV Sales (x1000)")
+
+cars_plot <- cars  %>% 
+	ggplot(aes(x = `Model Year`, y = `Production (000)`)) +
+	geom_point() +
+	ylim(c(0, NA)) +
+	labs(x = "Year",
+			 y = "Car Sales (x1000)")
+
+grid.arrange(ev_plot, cars_plot, ncol = 1)
+```
+
+And the share of EVs:
+
+```{r}
+evshare <- cars %>% 
+	transmute(Year = `Model Year`,
+						Cars = `Production (000)` * 1000) %>% 
+	left_join(evs, by = "Year") %>% 
+	mutate(ev_share = EVs / Cars)
+
+evshare %>% 
+	ggplot(aes(x = Year, y = ev_share)) +
+	geom_point() +
+	ylab("Share of EVs Sold vs. All Cars")
+```
+
+Finding a logistic growth model:
+
+```{r}
+data <- tibble(year = as.numeric(evshare$Year),
+							 share = as.numeric(evshare$ev_share))
+
+formula <- share~(exp(a+b*year)/(1+exp(a+b*year)))
+formula <- share~(exp(a)/(1+exp(a)))
+formula <- share~SSlogis(year, 1, median(year), 1)
+# model <- 
+	nls(formula,
+			# start = list(a = 1, b = 0.1),
+			data = data,
+			trace = TRUE)
+```

### 2022-09-21 HW2/HW2.Rmd
start HW2

--- 2022-09-09 HW1/HW1.Rmd
@@
 ---
-title: "HW 1"
+title: "HW 2: Energy Demand and CO2 Emissions"
 author: "Hayden Atchley"
-date: "9 September 2022"
+date: "`r Sys.Date()`"
 output: pdf_document
+mainfont: Gentium Book Basic
 ---
@@
 )
+
+library(tidyverse)
+library(gridExtra)
 ```
-
-# 1
-The systems boundary for SLC Int'l Airport is below:
-
-![SLC Int'l Airport systems boundary.](image/boundary.png)
-
-I think most of the elements of the diagram are fairly straightforward, where emissions sources produced directly by the airport are entirely inside the boundary, though a few elements are worth mentioning specifically.
-I considered personal travel to be a part of the trip in which a dropoff/pickup occurs, so the idling and minimal driving that happens with a dropoff/pickup is entirely within the airport boundary, while the transportation to and from the airport is not.
-Additionally, while the airport _does_ in a sense generate flights, the in-flight emissions themselves are outside the boundary of the airport (the emissions during landing, takeoff, and taxiing are entirely in the boundary).
-
-Another, perhaps easier, way to think about it (and how I thought about it) is a question of control.
-What emissions sources does the airport have total control over, and what sources are only partially under their control?
-The airport can, for example, control how much electricity it uses, but not the electricity to emissions ratio of the source.
-SLC International can (in theory at least) control the emissions of a plane while it's on the ground, but once the plane is in the air the airport has no control over the emissions.
-
-# 2
-The causal loop diagram is below:
-
-![Causal loop diagram exploring gas and electric vehicles.](image/loop.png)
-
-This analysis is far from perfect, but it gives a good idea of the relationships at play. I made a few assumptions:
-
- - An increase in gas price would lower the demand for (and therefore number of) gas-powered vehicles. This is probably not true in the short term, but is probably true in the long term.
- - Emissions regulations are likely to have an impact on fuel economy, though I'm not sure in which direction.
-Certain regulations may cause manufacturers to use engines that burn more cleanly, but that might have a negative impact on fuel economy.
-However, there may also be improvements in engine emissions that _increase_ fuel economy as well.
- - I am assuming that as the supply of rare earth metals decreases, battery technology will improve in a way that uses less of them. This may or not be the case, but I think it's likely.
-
-# 3

### 2022-09-21 HW2/HW2.Rmd
start hw@

@@
 date: "`r Sys.Date()`"
-output: pdf_document
+output:
+  pdf_document:
+    latex_engine: lualatex
 mainfont: Gentium Book Basic
@@
 library(gridExtra)
+library(kableExtra)
 ```
+
+# 1
+
+```{r}
+vm1 <- read_csv("HW2/data/vm1.csv", n_max = 28) %>% 
+	mutate(ITEM = ifelse(is.na(ITEM), lag(ITEM), ITEM))
+
+co2gal <- c(gasoline = 8887, diesel = 10180)
+```
+
+Creating a new category combining all buses, single-unit trucks, and combination trucks gives the following:
+
+```{r}
+hdv1 <- vm1 %>% 
+	mutate("ALL HDV" = BUSES + `SINGLE-UNIT TRUCKS/3` + `COMBINATION TRUCKS`)
+
+vmlong <- hdv1 %>% 
+	pivot_longer(c(-YEAR, -ITEM), names_to = "category") %>% 
+	arrange(category, ITEM)
+```
+
+## 1.1
+The average fuel economy is
+
+```{r}
+hdvfe <- hdv1 %>% 
+	filter(ITEM == "Average miles traveled per  gallon of fuel consumed") %>% 
+	select(YEAR, `ALL HDV`) %>% 
+	transmute(Year = YEAR, "Fuel economy (mpg)" = `ALL HDV`)
+
+hdvfe %>% 
+	kbl(booktabs = TRUE, caption = "Average Fuel Economy of HDVs") %>% 
+	kable_styling(latex_options = "HOLD_position")
+```
+
+## 1.2
+The CO~2~ emissions per mile is given by $\frac{1}{miles/gal} \times \frac{CO_2}{gal} = \frac{CO_2}{mile}$. This gives
+
+```{r}
+hdvem <- hdvfe %>% 
+	mutate("Fuel consumption (gpm)" = 1/`Fuel economy (mpg)`,
+				 "Emissions (g CO$_2$/gal)" = co2gal["diesel"],
+				 "Emissions (g CO$_2$/mile)" =
+				 	`Fuel consumption (gpm)` * `Emissions (g CO$_2$/gal)`)
+
+hdvem %>% 
+	kbl(booktabs = TRUE, escape = FALSE, caption = "HDV Fuel Consumption") %>% 
+	kable_styling(latex_options = "HOLD_position")
+```
+
+## 1.3
+The total CO~2~ emissions for all LDVs and HDVs is
+
+```{r}
+allem <- vmlong %>% 
+	filter(category %in% c("ALL LIGHT DUTY VEHICLES/2", "ALL HDV"),
+				 ITEM %in% c("Fuel consumed   (thousand gallons)")) %>% 
+	pivot_wider(names_from = ITEM) %>% 
+	mutate("Emissions (g CO$_2$/gal)" = c(rep(co2gal[2], 2),
+																				rep(co2gal[1], 2)),
+				 category = ifelse(str_detect(category, "LIGHT"),
+				 									"ALL LDV", category)) %>% 
+	arrange(desc(category), desc(YEAR)) %>% 
+	mutate("Emissions (g CO$_2$)" = 
+				 	`Fuel consumed   (thousand gallons)` *
+				 	`Emissions (g CO$_2$/gal)` * 1000,
+				 "Emissions (M tons CO$_2$)" = `Emissions (g CO$_2$/gal)` / 1000000)
+
+allem %>% 
+	kbl(booktabs = TRUE, escape = FALSE,
+			caption = "Emissions of LDVs and HDVs") %>% 
+	kable_styling(latex_options = c("HOLD_position", "scale_down"))
+```

### 2022-09-22 HW1/HW1.Rmd
finish hw2 1.3

@@
 ---
-title: "HW 1"
+title: "HW 1: System Tools"
 author: "Hayden Atchley"
-date: "9 September 2022"
+date: "`r Sys.Date()`"
 output: pdf_document
@@
 
-```{r}
+```{r, fig.height=2.7}
 evshare <- cars %>% 
@@
 
-Finding a logistic growth model:
+In order to solve this model, I first rearranged the growth equation:
+
+\[f(t) = \frac{e^{a+bt}}{1+e^{a+bt}} \implies
+f(t) + f(t)(e^{a+bt}) = e^{a+bt} \implies
+f(t) = e^{a+bt}(1-f(t)) \]
+\[\implies \frac{f(t)}{1-f(t)} = e^{a+bt} \implies
+\ln\left(\frac{f(t)}{1-f(t)}\right) = a+bt\]
+
+for all $0 < f(t) < 1$. \medskip
+
+Treating \(\ln\left(\frac{f(t)}{1-f(t)}\right)\) as the new dependent variable and plotting:
+
+```{r, fig.height=2.7}
+logevshare <- evshare %>% 
+	mutate(logshare = log(ev_share),
+				 Year = as.numeric(Year))
+
+fitline <- lm(logshare~Year, logevshare)
+
+logevshare %>% 
+	ggplot(aes(x = Year, y = logshare)) +
+	geom_point() +
+	geom_smooth(method = 'lm') +
+	scale_x_continuous(
+		breaks = min(logevshare$Year):max(logevshare$Year)) +
+	theme(panel.grid.minor.x = element_blank()) +
+	ylab("Log of EV Share")
+```
+
+Where the best-fit line is given by \(\ln\left(\frac{f(t)}{1-f(t)}\right) = -788.6 + 0.3889t,
+\ \therefore\  a=-788.6, b=0.3889\). Putting these values back in the original equation gives:
+
+\[f(t) = \frac{e^{-788.6 + 0.3889t}}{1+e^{-788.6 + 0.3889t}}.\]
+
+And adding this function to the original plot and expanding the time range:
 
 ```{r}
-data <- tibble(year = as.numeric(evshare$Year),
-							 share = as.numeric(evshare$ev_share))
+a <- fitline$coefficients["(Intercept)"]
+b <- fitline$coefficients["Year"]
 
-formula <- share~(exp(a+b*year)/(1+exp(a+b*year)))
-formula <- share~(exp(a)/(1+exp(a)))
-formula <- share~SSlogis(year, 1, median(year), 1)
-# model <- 
-	nls(formula,
-			# start = list(a = 1, b = 0.1),
-			data = data,
-			trace = TRUE)
+growthfun <- function(x){
+	exp(a+b*x)/(1+exp(a+b*x))
+}
+
+evshare %>%
+	mutate(Year = as.numeric(Year)) %>% 
+	ggplot(aes(x = Year, y = ev_share)) +
+	geom_point() +
+	geom_function(fun = growthfun) +
+	xlim(NA,2050) +
+	ylab("Share of EVs Sold vs. All Cars")
 ```
+
+```{r}
+time_at_share <- function(share){
+	(log(share/(1-share)) - a) / b
+}
+```
+
+Solving the equation
+\[0.5 = \frac{e^{-788.6 + 0.3889t}}{1+e^{-788.6 + 0.3889t}}\]
+gives a value of `r time_at_share(0.5) %>% round(1)`, which is the time at which EV sales account for half of all car sales.
+
+\newpage
+
+# 4
+The Annual Energy Outlook's estimation for EV penetration is given as:
+
+![Share of EV sales (AEO estimate).](image/AEO_estimate.png)
+
+This is a much shallower prediction, i.e. EVs do not penetrate nearly as quickly.

### 2022-09-22 HW2/HW2.Rmd
finish hw2 1.3

@@
 				 	`Emissions (g CO$_2$/gal)` * 1000,
-				 "Emissions (M tons CO$_2$)" = `Emissions (g CO$_2$/gal)` / 1000000)
+				 "Emissions (M tons CO$_2$)" = `Emissions (g CO$_2$)` / 1000000)
 
@@
 			caption = "Emissions of LDVs and HDVs") %>% 
+	collapse_rows(columns = 2, valign = "middle") %>% 
 	kable_styling(latex_options = c("HOLD_position", "scale_down"))
 ```
+
+Calculating the percentage of emissions from HDVs per year:
+
+```{r}
+pctem <- allem %>% 
+	select(YEAR, category, `Emissions (M tons CO$_2$)`) %>% 
+	pivot_wider(names_from = category, values_from = `Emissions (M tons CO$_2$)`) %>% 
+	transmute(Year = YEAR,
+						"LDV Emissions" = `ALL LDV`,
+						"HDV Emissions" = `ALL HDV`,
+						"HDV Proportion" = `HDV Emissions` /
+							(`HDV Emissions` + `LDV Emissions`))
+
+pctem %>% 
+	kbl(booktabs = TRUE, escape = FALSE,
+			caption = "HDV Emissions Proportion") %>% 
+	kable_styling(latex_options = c("HOLD_position", "scale_down"))
+```
+
+From this we can see that HDVs made up a greater percentage of CO~2~ emissions in 2020 than in 2019.
+
+## 1.4

### 2022-09-22 HW2/HW2.Rmd
update HW2

--- 2022-09-21 HW2/HW2.Rmd
@@
 				 	`Emissions (g CO$_2$/gal)` * 1000,
-				 "Emissions (M tons CO$_2$)" = `Emissions (g CO$_2$/gal)` / 1000000)
+				 "Emissions (M tons CO$_2$)" = `Emissions (g CO$_2$)` / 1000000) %>% 
+	relocate(category) %>% 
+	rename(Year = YEAR, "Vehicle category" = category)
 
@@
 			caption = "Emissions of LDVs and HDVs") %>% 
+	collapse_rows(columns = 1, valign = "middle") %>% 
 	kable_styling(latex_options = c("HOLD_position", "scale_down"))
 ```
+
+Calculating the percentage of emissions from HDVs per year:
+
+```{r}
+pctem <- allem %>% 
+	select(Year, `Vehicle category`, `Emissions (M tons CO$_2$)`) %>% 
+	pivot_wider(names_from = `Vehicle category`, values_from = `Emissions (M tons CO$_2$)`) %>% 
+	transmute(Year,
+						"LDV Emissions" = `ALL LDV`,
+						"HDV Emissions" = `ALL HDV`,
+						"HDV Proportion" = `HDV Emissions` /
+							(`HDV Emissions` + `LDV Emissions`))
+
+pctem %>% 
+	kbl(booktabs = TRUE, escape = FALSE,
+			caption = "HDV Emissions Proportion",
+			digits = 3) %>% 
+	kable_styling(latex_options = c("HOLD_position"))
+```
+
+From this we can see that HDVs made up a greater percentage of CO~2~ emissions in 2020 than in 2019.
+
+## 1.4
+
+Comparing the values I got to the values from the EPS's Greenhouse Gas Emissions Inventory:
+
+```{r}
+epaem <- read_csv("HW2/data/epa.csv") %>% 
+	filter(`U.S. Emissions, Transportation Sector, MMT CO2 eq.` == "Total") %>% 
+	select(`2019`:`2020`)
+
+epaemlist <- epaem %>% as.numeric()
+names(epaemlist) <- epaem %>% colnames()
+
+compareem <- pctem %>% 
+	select(-`HDV Proportion`) %>%
+	mutate("Total Emissions (MMT CO$_2$)" = 
+				 	(`LDV Emissions` + `HDV Emissions`) / 1000000) %>% 
+	tibble("EPA Values (MMT CO$_2$)" = rev(epaemlist))
+
+compareem %>% 
+	kbl(booktabs = TRUE, escape = FALSE,
+			caption = "Comparison to EPA Reported Emissions Values") %>% 
+	kable_styling(latex_options = c("HOLD_position"))
+```
+
+Though the values from my calculations differ from the EPA reported values, they are not wildly different from each other.
+They are close enough in my judgement that the difference could largely be explained by things like different measuring procedures.
+
+# 2
+
+The drive cycle I'm analyzing is the "Fleet DNA drayage Representative" cycle from NREL. The cycle is shown in Figure \ref{dc}.
+
+```{r, fig.height=4, fig.cap="Fleet DNA drayage Representative drive cycle.\\label{dc}", fig.pos="H"}
+
+dc <- read_csv("HW2/data/Fleet DNA Drayage Representative .csv")
+
+dc %>% 
+	ggplot(aes(x = `Time (seconds)`, y = `Speed (mph)`)) +
+	geom_line(color = "#004466") +
+	theme_minimal()
+```
+
+I am also using the assumed values in Table \ref{val} for my calculations.
+
+```{r}
+CD <- 0.6
+AV <- 8.5
+mE <- 11000
+mC <- 9000
+CR <- 0.006
+rhoa <- 1.17
+ag <- 9.81
+
+mv <- mE + mC
+
+mina <- -3.5
+maxa <- 2.5
+
+tribble(
+	~Coefficient, ~Symbol, ~Value,
+	"Drag coefficient", "$C_D$", "0.6",
+	"Frontal area", "$A_v$", "8.5 m$^2$",
+	"Empty vehicle weight", "$m_E$", "11,000 kg",
+	"Cargo weight", "$m_C$", "9,000 kg",
+	"Rolling resistance coefficient", "$C_R$", "0.006",
+	"Density of air", "$\\rho_a$", "1.17 kg/m$^3$",
+	"Acceleration of gravity", "$g$", "9.81 m/s$^2$"
+) %>% 
+	kbl(booktabs = TRUE, escape = FALSE,
+			caption = "Assumed Values for Calculations\\label{val}") %>% 
+	kable_styling(latex_options = c("HOLD_position"))
+```
+
+## 2.1
+The road-load equation for power is given by
+\[
+P_V = \frac{1}{2}\rho_aC_DA_vv^3 +
+v\left(C_Rm_vg + m_vg\sin\alpha + m_va_v\right),
+\]
+where $m_v = m_E + m_C$, and $\sin\alpha \approx \text{grade (rise/run)}$.
+
+Using this equation, we can calculate the tractive power for each second in the drive cycle.
+Note that since this is not a hybrid electric vehicle, there is no regenerative braking, and so all negative values of power are set to 0.
+We also set a limit on acceleration due to speed gaps causing unrealistic values.
+The limits are $`r mina` < a < `r maxa`$.
+
+```{r}
+tp <- dc %>% 
+	mutate(
+		time = `Time (seconds)`,
+		v = `Speed (mph)` * 0.44704,
+		araw = (v - lag(v)) / (time - lag(time)),
+		a = case_when(araw < mina ~ mina, araw > maxa ~ maxa, TRUE ~ araw),
+		grade = `Grade (rise/run)`,
+
+		dpowerraw = (1/2 * rhoa*CD*AV*v^3) / 1000,
+		dpower = ifelse(dpowerraw < 0, 0, dpowerraw),
+		rpowerraw = v*CR*mv*ag / 1000,
+		rpower = ifelse(rpowerraw < 0, 0, rpowerraw),
+		apowerraw = v*mv*a / 1000,
+		apower = ifelse(apowerraw < 0, 0, apowerraw),
+		gpowerraw = v*mv*ag*grade / 1000,
+		gpower = ifelse(gpowerraw < 0, 0, gpowerraw),
+
+		powerraw = dpowerraw + apowerraw + rpowerraw + gpowerraw,
+		power = ifelse(powerraw < 0, 0, powerraw)
+		) %>% 
+	select(-c(`Time (seconds)`, `Speed (mph)`, `Grade (rise/run)`)) %>% 
+	relocate(time, v)
+```
+
+The plot of the tractive power is given in Figure \ref{tp}.
+
+```{r, fig.cap="Plot of tractive power for drayage drive cycle.\\label{tp}", fig.height=4}
+tp %>% 
+	ggplot(aes(x = time, y = power)) +
+	geom_line(color = "#660044") +
+	theme_minimal() +
+	labs(x = "Time (seconds)",
+			 y = "Tractive Power (kW)")
+```
+
+## 2.2
+```{r}
+timemin <- 10000
+timemax <- 15000
+```
+
+A graph of engine power along with tractive power is given in Figure \ref{powercomp}.
+I am using only from `r timemin %>% format(scientific=FALSE)` to `r timemax %>% format(scientific=FALSE)` seconds for clarity of the graph.
+
+```{r, fig.height=4, fig.cap="Comparison of engine and tractive power.\\label{powercomp}"}
+powercomp <- tp %>% 
+	filter(time <= timemax, time >= timemin) %>% 
+	transmute(time, power, engine = `Engine Power (kW)`,
+						diff = power - engine) %>% 
+	pivot_longer(-time) %>% 
+	mutate(name = as.factor(name))
+
+powercomp$name <- powercomp$name %>%  
+	fct_relevel(c("power", "engine", "diff"))
+
+powercomp %>% 
+	ggplot(aes(x = time, y = value, color = name)) +
+	geom_line() +
+	labs(x = "Time (seconds)",
+			 y = "Power (kW)") +
+	scale_color_discrete(name = "", labels = c(
+		diff = "Engine - Tractive", engine = "Engine Power", power = "Tractive Power")) +
+	theme_minimal() + 
+  theme(legend.position = 'bottom')
+```
+
+The data generally seem to make sense, as the engine power is often more than the tractive power, which is expected due to losses.
+Though I am not sure why engine power sometimes exceeds tractive power; it appears to have something to do with the spikes in tractive power, which are perhaps due to some time gaps.
+
+## 2.3
+The total power required to move the vehicle for this drive cycle is `r sum(tp$power, na.rm=TRUE) %>% round() %>% format(scientific=FALSE)` kW.
+
+## 2.4
+```{r}
+drag <- sum(tp$dpower, na.rm = TRUE)
+roll <- sum(tp$rpower, na.rm = TRUE)
+accel <- sum(tp$apower, na.rm = TRUE)
+grav <- sum(tp$gpower, na.rm = TRUE)
+
+total <- sum(drag, roll, accel, grav)
+
+tibble(
+	" " = c("Power (kW)", "%"),
+	Drag = c(drag, drag/total * 100),
+	Rolling = c(roll, roll/total * 100),
+	Acceleration = c(accel, accel/total * 100),
+	Gravity = c(grav, grav/total * 100),
+	Total = c(total, total/total * 100)
+) %>% 
+	kbl(booktabs = TRUE, digits = 1, align = 'r',
+			caption = "Components of Total Tractive Power\\label{comp}") %>% 
+	kable_styling(latex_options = c("HOLD_position"))
+```
+
+## 2.5
+```{r}
+hours <- (max(tp$time) - min(tp$time)) / 3600
+
+regen <- -sum(tp$powerraw[tp$powerraw < 0], na.rm = TRUE)
+
+oldpower <- sum(tp$power, na.rm = TRUE)
+newpower <- oldpower - regen
+```
+
+Assuming the "negative" tractive power would be used for regenerative braking, an additional `r (regen*hours) %>% round() %>% format(scientific=FALSE)` kWh would be available (`r regen %>% round() %>% format(scientific=FALSE)` Kw over `r hours %>% round(2)` hours).
+If 50% of that power was recaptured, then the additional tractive power needed would be reduced to `r newpower %>% round() %>% format(scientific=FALSE)` kW, a reduction of `r ((oldpower - newpower)/oldpower * 100) %>% round(2)`%.

### 2022-09-28 HW2/HW2.Rmd
update hw2

@@
 library(kableExtra)
+
+options(scipen = 999)
 ```
@@
 		powerraw = dpowerraw + apowerraw + rpowerraw + gpowerraw,
-		power = ifelse(powerraw < 0, 0, powerraw)
+		power = ifelse(powerraw < 0, 0, powerraw),
+
+		dist = v*(time - lag(time))
 		) %>% 
@@
 
-A graph of engine power along with tractive power is given in Figure \ref{powercomp}.
-I am using only from `r timemin %>% format(scientific=FALSE)` to `r timemax %>% format(scientific=FALSE)` seconds for clarity of the graph.
+A graph of engine power along with tractive power is given in Figure \ref{powercomp}. I am using only from `r timemin` to `r timemax` seconds for clarity of the graph.
 
@@
 	transmute(time, power, engine = `Engine Power (kW)`,
-						diff = power - engine) %>% 
+						diff = engine - power) %>% 
 	pivot_longer(-time) %>% 
@@
 	scale_color_discrete(name = "", labels = c(
-		diff = "Engine - Tractive", engine = "Engine Power", power = "Tractive Power")) +
+		diff = "Tractive minus Engine", engine = "Engine Power", power = "Tractive Power")) +
 	theme_minimal() + 
@@
 The data generally seem to make sense, as the engine power is often more than the tractive power, which is expected due to losses.
-Though I am not sure why engine power sometimes exceeds tractive power; it appears to have something to do with the spikes in tractive power, which are perhaps due to some time gaps.
+Though I am not sure why tractive power sometimes exceeds engine power; it appears to have something to do with the spikes in tractive power, which are perhaps due to some time gaps.
 
 ## 2.3
-The total power required to move the vehicle for this drive cycle is `r sum(tp$power, na.rm=TRUE) %>% round() %>% format(scientific=FALSE)` kW.
+The total power required to move the vehicle for this drive cycle is `r sum(tp$power, na.rm=TRUE) %>% round()` kW.
 
@@
 
-Assuming the "negative" tractive power would be used for regenerative braking, an additional `r (regen*hours) %>% round() %>% format(scientific=FALSE)` kWh would be available (`r regen %>% round() %>% format(scientific=FALSE)` Kw over `r hours %>% round(2)` hours).
-If 50% of that power was recaptured, then the additional tractive power needed would be reduced to `r newpower %>% round() %>% format(scientific=FALSE)` kW, a reduction of `r ((oldpower - newpower)/oldpower * 100) %>% round(2)`%.
+Assuming the "negative" tractive power would be used for regenerative braking, an additional `r (regen*hours) %>% round()` kWh would be available (`r regen %>% round()` kW over `r hours %>% round(2)` hours).
+If 50% of that power was recaptured, then the additional tractive power needed would be reduced to `r newpower %>% round()` kW, a reduction of `r ((oldpower - newpower)/oldpower * 100) %>% round(2)`%.
+
+## 2.6
+```{r}
+enginemove <- tp %>% 
+	mutate(newengine = `Engine Power (kW)` - 7,
+				 newengine = ifelse(newengine < 0, 0, newengine))
+
+moveenergy <- sum(enginemove$newengine, na.rm = TRUE)*hours
+```
+
+Assuming an auxiliary power load of 7 kW, the engine power used to move the vehicle is $P_{Engine} - 7$ kW, but not less than 0 kW.
+Finding the total energy expended by the engine attempting to move the vehicle gives `r moveenergy %>% round()` kWh.
+The total energy actually required to move the vehicle is `r (oldpower*hours) %>% round()` kWh, so the efficiency of the drivetrain is `r (oldpower*hours/moveenergy * 100) %>% round(1)`%.
+
+## 2.7
+Low-sulfur diesel has an energy content of 128,488 Btu/gal according to the AFDC.
+This is equivalent to `r (energy <- 37.656)` kWh/gal.
+Assuming an engine efficiency of `r (eff <- 0.3)`, the total fuel needed for this drive cycle is $F[\text{gal}] = \frac{`r (oldpower*hours) %>% round()`[\text{kWh}]}{`r energy`[\text{kWh/gal}]}/`r eff` = `r (fuel <- oldpower*hours/energy/eff) %>% round()`\ \text{gal}$.
+The total distance in this drive cycle is `r (totdist <- sum(tp$dist, na.rm = TRUE)) %>% round()` meters or `r (totdist * .0006213712) %>% round(1)` miles, giving

### 2022-09-28 HW2/HW2.Rmd
finish 2:2

@@
 	kbl(booktabs = TRUE, escape = FALSE,
-			caption = "Assumed Values for Calculations\\label{val}") %>% 
+			caption = "Assumed Values for Calculations\\label{val}",
+			linesep = "") %>% 
 	kable_styling(latex_options = c("HOLD_position"))
@@
 		time = `Time (seconds)`,
+		t = time - lag(time),
 		v = `Speed (mph)` * 0.44704,
-		araw = (v - lag(v)) / (time - lag(time)),
+		araw = (v - lag(v)) / t,
 		a = case_when(araw < mina ~ mina, araw > maxa ~ maxa, TRUE ~ araw),
@@
 
-		dist = v*(time - lag(time))
+		energy = power*t/3600, #kWh
+
+		dist = v*t
 		) %>% 
@@
 ```{r}
-timemin <- 10000
-timemax <- 15000
+timemin <- 13900
+timemax <- 14000
 ```
@@
 	transmute(time, power, engine = `Engine Power (kW)`,
-						diff = engine - power) %>% 
+						diff = power - engine) %>% 
 	pivot_longer(-time) %>% 
@@
 The data generally seem to make sense, as the engine power is often more than the tractive power, which is expected due to losses.
-Though I am not sure why tractive power sometimes exceeds engine power; it appears to have something to do with the spikes in tractive power, which are perhaps due to some time gaps.
+Though I am not sure why tractive power sometimes exceeds engine power; it appears to have something to do with the spikes in tractive power, which might be due to some time gaps in the data, or possibly a small temporal offset between the engine and tractive power.
 
 ## 2.3
-The total power required to move the vehicle for this drive cycle is `r sum(tp$power, na.rm=TRUE) %>% round()` kW.
+The total energy required to move the vehicle for this drive cycle is `r sum(tp$energy, na.rm=TRUE) %>% round()` kWh.
 
@@
 ```{r}
-drag <- sum(tp$dpower, na.rm = TRUE)
-roll <- sum(tp$rpower, na.rm = TRUE)
-accel <- sum(tp$apower, na.rm = TRUE)
-grav <- sum(tp$gpower, na.rm = TRUE)
+drag <- sum(tp$dpower*tp$t/3600, na.rm = TRUE)
+roll <- sum(tp$rpower*tp$t/3600, na.rm = TRUE)
+accel <- sum(tp$apower*tp$t/3600, na.rm = TRUE)
+grav <- sum(tp$gpower*tp$t/3600, na.rm = TRUE)
 
@@
 tibble(
-	" " = c("Power (kW)", "%"),
+	" " = c("Energy (kWh)", "%"),
 	Drag = c(drag, drag/total * 100),
@@
 ```{r}
-hours <- (max(tp$time) - min(tp$time)) / 3600
+regen <- -sum(tp$powerraw[tp$powerraw < 0] * 
+								tp$t[tp$powerraw < 0]/3600, na.rm = TRUE)
 
-regen <- -sum(tp$powerraw[tp$powerraw < 0], na.rm = TRUE)
-
-oldpower <- sum(tp$power, na.rm = TRUE)
-newpower <- oldpower - regen
+oldpower <- sum(tp$energy, na.rm = TRUE)
+newpower <- oldpower - (regen/2)
 ```
 
-Assuming the "negative" tractive power would be used for regenerative braking, an additional `r (regen*hours) %>% round()` kWh would be available (`r regen %>% round()` kW over `r hours %>% round(2)` hours).
-If 50% of that power was recaptured, then the additional tractive power needed would be reduced to `r newpower %>% round()` kW, a reduction of `r ((oldpower - newpower)/oldpower * 100) %>% round(2)`%.
+Assuming the "negative" tractive power would be used for regenerative braking, an additional `r (regen) %>% round()` kWh would be available.
+If 50% of that power was recaptured, then the additional tractive power needed would be reduced to `r newpower %>% round()` kWh, a reduction of `r ((oldpower - newpower)/oldpower * 100) %>% round(2)`%.
 
@@
 
-moveenergy <- sum(enginemove$newengine, na.rm = TRUE)*hours
+moveenergy <- sum(enginemove$newengine*enginemove$t/3600, na.rm = TRUE)
 ```
@@
 Finding the total energy expended by the engine attempting to move the vehicle gives `r moveenergy %>% round()` kWh.
-The total energy actually required to move the vehicle is `r (oldpower*hours) %>% round()` kWh, so the efficiency of the drivetrain is `r (oldpower*hours/moveenergy * 100) %>% round(1)`%.
+The total energy actually required to move the vehicle is `r (oldpower) %>% round()` kWh, so the efficiency of the drivetrain is `r (oldpower/moveenergy * 100) %>% round(1)`%.
 
@@
 This is equivalent to `r (energy <- 37.656)` kWh/gal.
-Assuming an engine efficiency of `r (eff <- 0.3)`, the total fuel needed for this drive cycle is $F[\text{gal}] = \frac{`r (oldpower*hours) %>% round()`[\text{kWh}]}{`r energy`[\text{kWh/gal}]}/`r eff` = `r (fuel <- oldpower*hours/energy/eff) %>% round()`\ \text{gal}$.
-The total distance in this drive cycle is `r (totdist <- sum(tp$dist, na.rm = TRUE)) %>% round()` meters or `r (totdist * .0006213712) %>% round(1)` miles, giving
+Assuming an engine efficiency of `r (eff <- 0.3)`, the total fuel needed for this drive cycle is $F = \frac{`r (oldpower) %>% round()`[\text{kWh}]/`r eff`}{`r energy %>% round(1)`[\text{kWh/gal}]} = `r (fuel1 <- oldpower/energy/eff) %>% round()`\ \text{gal}$.
+The total distance in this drive cycle is `r (totdist <- sum(tp$dist, na.rm = TRUE)) %>% round()` meters or `r (miles <- (totdist * .0006213712)) %>% round(1)` miles, giving an overall fuel economy of `r (miles/fuel1) %>% round(2)` mpg, and a fuel consumption of `r (fuel1/miles*1000) %>% round()` gallons per 1000 miles.
+
+## 2.8
+The energy required in the regenerative-braking scenario (2.5) is `r newpower %>% round()` kWh.
+The fuel needed is now $F = \frac{`r (newpower) %>% round()`[\text{kWh}]/`r eff`}{`r energy %>% round(1)`[\text{kWh/gal}]} = `r (fuel2 <- newpower/energy/eff) %>% round()`\ \text{gal}$, with a fuel economy and consumption of `r (miles/fuel2) %>% round(2)` mpg and `r (fuel2/miles*1000) %>% round()` gallons per 1000 miles.
+If this drive cycle (`r (miles <- (totdist * .0006213712)) %>% round(1)` miles) is representative of a typical day, then at 5 days per week and 52 weeks per year, the fuel saved is $5 \text{[days/wk]} \times 52 \text{[wk/yr]} \times (`r fuel1 %>% round()` - `r fuel2 %>% round()`) \text{[gal]} = `r (fuelsave <- 5*52*(fuel1-fuel2)) %>% round` \text{[gal/yr]}$ for each truck.
+Given a current (26 Sept.
+2022) diesel price of \$`r (price <- 4.889)`/gal, this amounts to a savings of \$`r (price*fuelsave) %>% round()` per year per truck.

### 2022-09-30 HW2/HW2.Rmd
finish? hw2

@@
   pdf_document:
-    latex_engine: lualatex
+    latex_engine: xelatex
 mainfont: Gentium Book Basic
@@
 2022) diesel price of \$`r (price <- 4.889)`/gal, this amounts to a savings of \$`r (price*fuelsave) %>% round()` per year per truck.
+
+# 3
+From Annual Energy Outlook 2022 Table 40, we get that the 2021 fuel economies of certain vehicle technologies are as follows:
+
+```{r}
+gasmpg <- 42.94
+hybmpg <- 62.51
+evmpg <- 101.9
+
+kwhtogal <- 0.03
+
+coalem <- 767/757763 * 1000
+ngem <- 576/1402438 * 1000
+```
+
+- ICE vehicles: `r gasmpg` mpg
+- Hybrid vehicles: `r hybmpg` mpge
+- 200 mile range EVs: `r evmpg` mpge
+
+From the EIA we obtain the emission rate of coal and natural gas energy production:
+
+```{r}
+emissions <- tibble(
+	"Energy Source" = c("Coal", "Natural Gas", "Petroleum"),
+	electricity = c(757763, 1402438, 13665),
+	co2 = c(767, 576, 13),
+	emrate = co2 / electricity * 1000
+	)
+
+emissions %>% 
+	rename(
+		"Electricity (M kWh)" = electricity,
+		"CO$_2$ Emissions (M MT)" = co2,
+		"CO$_2$ per kWh (kg)" = emrate
+	) %>% 
+	kbl(digits = 3,
+			caption = "CO$_2$ Emissions per Kilowatt-hour Produced",
+			escape = FALSE,
+			booktabs = TRUE) %>% 
+	kable_styling(latex_options = c("HOLD_position"))
+
+emc <- emissions$co2[1]
+emng <- emissions$co2[2]
+
+elecc <- emissions$electricity[1]
+elecng <- emissions$electricity[2]
+
+emratep <- emissions$emrate[3]
+
+weightem <- ((emc + emng) / (elecc + elecng)) * 1000
+```
+
+Assuming the electricity used in the EVs is produced from coal and natural gas, we can calculate an average of the two emission rates.
+By summing the total emissions and total energy generated, we find that the overall emission rate is $\frac{`r emc` + `r emng`}{`r elecc` + `r elecng`} \times 1000 = `r weightem %>% round(2)`$ kg CO~2~ per kWh.
+
+## 3.1
+The following table shows information on CO~2~ emissions by vehicle type.
+I am assuming that the emission rate from petroleum electricity generation is similar enough to the rate when using an ICE vehicle to use in these calculations.
+I am also using a conversion factor of 1 kWh to `r kwhtogal` gallons of gasoline.
+
+```{r}
+emtable <- tibble(
+	"Vehicle Type" = c("ICE", "Hybrid", "EV"),
+	mpg = c(gasmpg, hybmpg, evmpg),
+	mpkwh = mpg*kwhtogal,
+	co2pkwh = c(emratep, emratep, weightem),
+	co2pmi = co2pkwh / mpkwh * 1000
+)
+
+emtable %>% 
+	rename(
+		"MPG (equiv)" = mpg,
+		"Miles per kWh" = mpkwh,
+		"CO$_2$ per kWh (kg)" = co2pkwh,
+		"CO$_2$ per Mile (g)" = co2pmi
+	) %>% 
+	kbl(digits = c(0,1,2,3,0),
+			caption = "CO$_2$ Emissions per Mile by Vehicle Type",
+			escape = FALSE,
+			booktabs = TRUE) %>% 
+	kable_styling(latex_options = c("HOLD_position"))
+```
+
+The EV has the lowest emission rate per mile, while the ICE vehicle has the highest.
+This follows my intuition, as even with the inefficiencies in electricity production EVs are much more efficient per kWh than gas-powered vehicles.
+The emission rate of the EVs could theoretically fall even further if cleaner electricity sources are used.
+
+## 3.2
+Using the EPA 2020 average of 25.5 mpg and 350 g CO~2~/mile for the ICE vehicle, the hybrid vehicle now has the worst emission rate.
+Though in reality this probably would not be the case; a lower emission rate per gallon of gasoline used would benefit both ICE and hybrid vehicles.
+It's possible my previous assumption that emissions would be similar between an ICE and petroleum electricity generation is misguided, in which case the numbers I calculated may change significantly.

### 2022-10-14 project/proposal.Rmd
start project proposal

---
title: Project Proposal
author: Hayden Atchley
date: '`r format(Sys.Date(), "%d %B %Y")`'
output:
  bookdown::pdf_document2:
    latex_engine: lualatex
    toc: false
    number_sections: false
mainfont: Gentium Book Basic
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```

# Introduction

Recently, electric vehicles (EVs) have gained significant public awareness and various levels of support. Generally the adoption of EVs is seen as a good thing, as there are many clear benefits over internal combustion engine (ICE) vehicles. Perhaps the most obvious is that EVs have no emissions, since they are driven entirely by electric motors. Another important advantage is efficiency: while an ICE vehicle will convert around ####% of the fuel's stored energy into motion, EVs are nearly 100% efficient. Electric motors are also efficient over a large range of rotational speeds, whereas ICEs are only efficient from roughly ### to ### rpm.

However, EVs are not always universally superior to ICE vehicles. One major consideration is that the range of EVs is lower than may be practical in certain applications. Though EVs are more efficient at converting stored energy into motion, gasoline and diesel are more energy-dense than batteries, and so that energy becomes harder to store in an EV by comparison. What truly makes this a concern is the time required to recharge an EV's battery compared to the time to refuel an ICE vehicle's tank. Though there are relatively fast charging standards available, as of now none of them are nearly as fast as filling up a car at a gas station.

This charging problem, though, is not generally a concern for personal vehicles. Even the slowest of charging standards are able to charge at ####W, which if charged overnight could easily offset a daily commute of ### miles. However, for one major part of the transportation industry, namely commercial transportation, EV charging becomes a major consideration. Long-haul trucks, bus rapid transit (BRT), and even local bus networks may not be as well-suited for EV adoption.

# Research Question

So how can a local transit agency decide if electrification of their bus network is a worthwhile pursuit? What are the factors that must be considered, and what are some potential problems and/or solutions? What are the costs of these solutions, and are they practical?

I aim to analyze these questions, with a specific consideration of Utah Transit Authority's (UTA's) bus network, including the Utah Valley eXpress (UVX) BRT as well as the local bus network in the Provo/Orem area. Ultimately I plan to determine what would be required to electrify this bus system and if it is feasible for UTA to do so.

I also hope to provide somewhat of a framework to help in making similar decisions in the future, both for UTA and other agencies. Recognizing these required considerations and planning for them is an important step in making future electrification easier and more worthwhile.

# Preliminary Literature Review

# What methods/data will I use to answer the question?

I will

### 2022-10-15 project/proposal.Rmd
finish project proposal

@@
 Perhaps the most obvious is that EVs have no emissions, since they are driven entirely by electric motors.
-Another important advantage is efficiency: while an ICE vehicle will convert around ####% of the fuel's stored energy into motion, EVs are nearly 100% efficient.
-Electric motors are also efficient over a large range of rotational speeds, whereas ICEs are only efficient from roughly ### to ### rpm.
+Another important advantage is efficiency: while an ICE vehicle will convert around 20% of the fuel's stored energy into motion [@gas-efficiency], EVs are around 90% efficient [@ev-efficiency].
+Electric motors are also efficient over a large range of rotational speeds, whereas ICEs are only efficient at a much narrower range of speeds.
 
@@
 What truly makes this a concern is the time required to recharge an EV's battery compared to the time to refuel an ICE vehicle's tank.
-Though there are relatively fast charging standards available, as of now none of them are nearly as fast as filling up a car at a gas station.
+Though there are relatively fast charging standards available, as of now none of them are nearly as fast as filling up a car at a gas station [@epa-charging].
 
 This charging problem, though, is not generally a concern for personal vehicles.
-Even the slowest of charging standards are able to charge at ####W, which if charged overnight could easily offset a daily commute of ### miles.
+Even the slowest of charging standards are able to charge at up to 1.9kW, which if charged either overnight or at a workplace for 8 hours could easily offset a daily commute of 40 miles or more [@epa-charging].
 However, for one major part of the transportation industry, namely commercial transportation, EV charging becomes a major consideration.
@@
 
+Developing charging infrastructure is the most significant prerequisite to deploying a fleet of EV buses.
+There are currently several options available for EV charging.
+Level 1 and 2 charging as well as DC fast charging use a plug-in connector to transfer power [@epa-charging], and other options such as wireless inductive charging exist and are being developed upon [@Klontz1993; @Panchal2018].
+I aim to analyze which, if any, of these options would be appropriate.
+
 I also hope to provide somewhat of a framework to help in making similar decisions in the future, both for UTA and other agencies.
@@
 
-# Preliminary Literature Review
+# Planned Methodology
 
-# What methods/data will I use to answer the question?
+In order to perform this analysis, I will also need to obtain information on UTA's bus fleets.
+This will include current vehicle types, fleet sizes, routes, and operating schedules.
+Currently, the UVX service uses hybrid electric vehicles, but this may not be true for all of the local bus routes.
+The size and weight of the vehicles is also an important consideration, as that will at least somewhat affect the size of battery pack that can be used, and therefore the range of the vehicle.
 
-I will
+I will continue to study the literature surrounding EV charging, and using this information as well as the information about UTA's bus fleets I plan to determine (a) possible path(s) to electrification of the bus service.
+I will analyze the cost of this(ese) path(s) and suggest a potential "best solution".
+
+# References

### 2022-11-08 HW4/HW4.Rmd
start hw4

---
title: "HW #4: Analyze real-world emissions data"
author: Hayden Atchley
date: "`r Sys.Date()`"
mainfont: Gentium Book Basic
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)

pacman::p_load(tidyverse, kableExtra, zfit, broom, ggthemes, DescTools)
```

```{r}
feat <- read_csv("data/feat.class.example.csv")
```

# 1

```{r}
#| em-density,
#| fig.cap="Emissions density by date.",
#| fig.pos="H"

feat1 <- feat %>% 
	replace_na(list(NO_g_kg = 0, NO2_g_kg = 0, HC_g_kg = 0, CO_g_kg = 0)) %>% 
	mutate(NOx_g_kg = NO_g_kg + NO2_g_kg) %>% 
	select(DATE, CO_g_kg, HC_g_kg, NOx_g_kg, location, VEH_ID) %>% 
	pivot_longer(-c(DATE, location, VEH_ID), names_to = "pollutant", values_to = "amount") %>% 
	mutate(pollutant = str_remove(pollutant, "_.+"))

feat1 %>% 
	ggplot() +
	geom_density(aes(x = amount, color = pollutant)) +
	facet_wrap(~DATE) +
	lims(x = c(NA,200),
			 y = c(NA,0.1)) +
	labs(x = "Emmission Rate (g / kg fuel)",
			 y = "kernel density",
			 color = "Pollutant")
```

## 1.1

```{r}
feat1 %>% 
	group_by(location, pollutant) %>% 
	summarise(max_emisions = max(amount), median_emissions = median(amount)) %>% 
	mutate("max / median" = max_emisions / median_emissions) %>% 
	kbl(booktabs = TRUE, digits = 2, linesep = "") %>% 
	kable_styling(latex_options = "HOLD_position")
```

## 1.2

Looking at Figure \@ref(fig:em-density), NOx appears to have quite a skewed distribution, though CO has a few extreme outliers.

# 2

```{r}
#| fig.height=3

feat2 <- feat1 %>% 
	select(VEH_ID, pollutant, amount) %>% 
	pivot_wider(names_from = pollutant, values_from = amount)

feat2 %>% 
	ggplot(aes(x = CO, y = HC)) +
	geom_point() +
	geom_smooth(method = 'lm') +
	theme_pander()
```

```{r}
#| fig.height=3

feat2 %>% 
	ggplot(aes(x = CO, y = NOx)) +
	geom_point() +
	geom_smooth(method = 'lm') +
	theme_pander()
```

```{r}
fit2 <- feat2 %>% 
	zlm(CO ~ NOx + HC) %>% 
	summary()

fit2 %>% 
	tidy() %>% 
	kbl(booktabs = TRUE, digits = c(0,2,2,2,3), escape = FALSE) %>% 
	footnote(paste0("$R^2$ = ", fit2$adj.r.squared %>% round(3)),
					 general_title = "") %>% 
	kable_styling(latex_options = "HOLD_position")
```

While it appears that vehicles with more CO emissions also have more NOx and HC emissions (both of these slopes/coefficients are positive), the $R^2$ value is quite low. There could be many other factors explaining the variance in emission rates.

# 3

# 4

```{r}
feat1 %>% 
	filter(pollutant == "CO") %>% 
	select(-pollutant, -VEH_ID) %>% 
	group_by(DATE, location) %>% 
	summarise(
		name = names(MeanCI(amount, conf.level = 0.9)),
		value = MeanCI(amount, conf.level = 0.9)) %>% 
	pivot_wider()
```

```{r}
feat1 %>% 
	filter(pollutant == "CO") %>% 
	select(-pollutant, -VEH_ID) %>% 
	ggplot(aes(y = DATE, x = amount, fill = location)) +
	geom_violin() +
	stat_summary(
		fun.data = function(x) MeanCI(x, conf.level = 0.9) %>% 
			`names<-`(c("y", "ymin", "ymax")),
		geom = "errorbar",
		width = 0.3) +
	stat_summary(fun = mean, geom = "point")
```

### 2022-11-18 HW4/HW4.Rmd
update hw4

@@
 	facet_wrap(~DATE) +
-	lims(x = c(NA,200),
-			 y = c(NA,0.1)) +
+	theme_pander() +
+	lims(x = c(NA,100),
+			 y = c(NA, 0.3)) +
 	labs(x = "Emmission Rate (g / kg fuel)",
@@
 	tidy() %>% 
-	kbl(booktabs = TRUE, digits = c(0,2,2,2,3), escape = FALSE) %>% 
+	kbl(booktabs = TRUE, digits = c(0,2,2,2,3)) %>% 
 	footnote(paste0("$R^2$ = ", fit2$adj.r.squared %>% round(3)),
-					 general_title = "") %>% 
+					 general_title = "",
+					 escape = FALSE) %>% 
 	kable_styling(latex_options = "HOLD_position")
@@
 		value = MeanCI(amount, conf.level = 0.9)) %>% 
-	pivot_wider()
+	pivot_wider() %>% 
+	kbl(booktabs = TRUE, digits = c(0,0,2,2,1)) %>% 
+	kable_styling(latex_options = "HOLD_position")
 ```
@@
 ```{r}
+#| mean-conf,
+#| fig.cap = "CO emission rate for each day. The bars show the mean and 90% confidence interval for the mean.",
+#| fig.pos = "H",
+#| fig.height = 3
+
 feat1 %>% 
@@
 		width = 0.3) +
-	stat_summary(fun = mean, geom = "point")
+	stat_summary(fun = mean, geom = "point") +
+	theme_pander() +
+	xlim(NA, 200) +
+	labs(
+		x = "CO Emission Rate (g / kg fuel)",
+		y = "Date",
+		fill = "Location")
 ```
+
+```{r}
+feat1 %>% 
+	filter(pollutant == "CO") %>% 
+	aov(amount ~ location, data = .) %>% 
+	TukeyHSD()
+```

### 2022-11-19 HW4/HW4.Rmd
almost finish hw4

@@
 
+The table below shows a Tukey-Kramer test for multiple variance based on emissions test location.
+None of the _p_-values indicate significant results, so we can't conclude that the means are different from each other.
+
 ```{r}
@@
 	aov(amount ~ location, data = .) %>% 
-	TukeyHSD()
+	TukeyHSD() %>% 
+	tidy() %>% 
+	select(-term, -null.value)
 ```
+
+## 4.1
+
+Assuming the re-sampling method refers to a permutation/randomization test, the main advantage is that no assumptions need to be made regarding skewness, normality, or outliers.
+The main disadvantage is that it is computationally intensive to compare all permutations of the data points, especially when the data sets are large.
+Often a smaller subset of permutations is performed, which can usually offer a good approximation of the full permutation test.
+
+# 5
+
+# 6
+
+```{r}
+site <- read_csv("data/CO.site.average.date.2023.csv") %>% 
+	mutate(DATE = as.Date(DATE, format = "%m/%d/%Y"))
+```
+
+Looking at a plot of the emission rates over time, there doesn't appear to be any meaningful trend:
+
+```{r}
+#| fig.pos = "H",
+#| fig.height = 3
+
+site %>% 
+	ggplot(aes(x = DATE, y = mean_CO_g_kg)) +
+	geom_point() +
+	theme_pander() +
+	labs(
+		x = "Date",
+		y = "Mean CO Emissions (g / kg fuel)"
+	)
+```
+
+Calculating the mean of these values with a 95% confidence interval gives:
+
+```{r}
+MeanCI(site$mean_CO_g_kg) %>% 
+	bind_rows() %>% 
+	kbl(digits = 1, booktabs = TRUE) %>% 
+	kable_styling(latex_options = "HOLD_position")
+```

### 2022-11-23 project/Paper.Rmd
start paper draft

--- 2022-10-15 project/proposal.Rmd
@@
 ---
-title: Project Proposal
+title: Electrification of Bus Networks in Provo, Utah (DRAFT)
 author: Hayden Atchley
-date: '`r format(Sys.Date(), "%d %B %Y")`'
+date: "22 November 2022"
+documentclass: article
+geometry:
+  - letterpaper
+  - margin=1in
 output:
@@
     number_sections: false
+header-includes:
+  - \usepackage{setspace}
 mainfont: Gentium Book Basic
@@
 ---
+
+\doublespace
 
@@
 Long-haul trucks, bus rapid transit (BRT), and even local bus networks may not be as well-suited for EV adoption.
+But despite the difficulty of solving the charging problem, electrification of these systems can still be a worthwhile goal, due to the many advantages of EVs.
+In fact, many transit agencies are using electric bus fleets XXXXXXXX, and others are in the process of transitioning to EVs XXXXXXX.
 
-# Research Question
-
-So how can a local transit agency decide if electrification of their bus network is a worthwhile pursuit?
-What are the factors that must be considered, and what are some potential problems and/or solutions?
-What are the costs of these solutions, and are they practical?
-
-I aim to analyze these questions, with a specific consideration of Utah Transit Authority's (UTA's) bus network, including the Utah Valley eXpress (UVX) BRT as well as the local bus network in the Provo/Orem area.
-Ultimately I plan to determine what would be required to electrify this bus system and if it is feasible for UTA to do so.
+# Charging Solutions
 
@@
 Level 1 and 2 charging as well as DC fast charging use a plug-in connector to transfer power [@epa-charging], and other options such as wireless inductive charging exist and are being developed upon [@Klontz1993; @Panchal2018].
-I aim to analyze which, if any, of these options would be appropriate.
 
-I also hope to provide somewhat of a framework to help in making similar decisions in the future, both for UTA and other agencies.
-Recognizing these required considerations and planning for them is an important step in making future electrification easier and more worthwhile.
-
-# Planned Methodology
-
-In order to perform this analysis, I will also need to obtain information on UTA's bus fleets.
-This will include current vehicle types, fleet sizes, routes, and operating schedules.
-Currently, the UVX service uses hybrid electric vehicles, but this may not be true for all of the local bus routes.
-The size and weight of the vehicles is also an important consideration, as that will at least somewhat affect the size of battery pack that can be used, and therefore the range of the vehicle.
-
-I will continue to study the literature surrounding EV charging, and using this information as well as the information about UTA's bus fleets I plan to determine (a) possible path(s) to electrification of the bus service.
-I will analyze the cost of this(ese) path(s) and suggest a potential "best solution".
+# Research Question
 

### 2022-11-23 project/Paper.Rmd
update paper

@@
 Level 1 and 2 charging as well as DC fast charging use a plug-in connector to transfer power [@epa-charging], and other options such as wireless inductive charging exist and are being developed upon [@Klontz1993; @Panchal2018].
+Perhaps somewhat unique to buses are <><><fancy overhead chargers><><>, where a rail on the bus extends to make contact with an overhead power supply at a bus stop.
 
-# Research Question
+While the level 1 and level 2 EV charging standards work well for personal vehicles, which sit idle for large parts of the day, these are relatively slow standards.
+A personal vehicle can expect about 25 miles of range per hour of level 2 charging [@epa-charging].
+A bus, which weighs several times that of a personal vehicle [@car-weights], would get a fraction of that range (since work/energy is proportional to mass).
+Many bus routes run for most of the day, and even with an 8-hour daily level 2 charge a bus may only gain about 50 miles of daily range.
+A city bus running for 12 hours per day with an average speed of 12mph would need more than double that.
+Exclusively using level 2 charging would in many cases necessitate at least a doubling of fleet size to maintain those routes, which is impractical.
+
+DC fast charging can charge vehicles on the order of 10 times as fast as level 2 charging [@epa-charging].
+Depending on the implementation, this standard can charge as fast as 350kW.
+This is certainly fast enough to fully recharge an electric bus overnight, so seems to be a viable way to solve the charging problem.
+However, standard plug-in DC fast charging overnight may not be enough on its own due to battery sizes.
+
+EV range is directly a function of battery pack size, and so a larger pack will provide more range.
+A typical electric bus with a battery pack size of 350kWh and energy consumption of about 3kWh per mile would have around 100 miles of range [@ebus-range].
+This may be sufficient for many routes much of the year, but electric heating in the winter months can drastically reduce EV range.
+Assuming a worst-case scenario of 4-6kWh per mile, this reduces the range of a 350kWh bus to around 70 miles.
+While some buses are being developed with larger battery packs [@big-ebus], these are currently not commonly in use.
 

### 2022-12-02 HW4/HW4.Rmd
redo hw4.6

@@
 site <- read_csv("data/CO.site.average.date.2023.csv") %>% 
-	mutate(DATE = as.Date(DATE, format = "%m/%d/%Y"))
+	mutate(DATE = as.Date(DATE, format = "%m/%d/%Y")) %>% 
+	group_by(location) %>% 
+	summarise(mean_loc = mean(mean_CO_g_kg))
 ```
 
-Looking at a plot of the emission rates over time, there doesn't appear to be any meaningful trend:
+Because there were multiple data points at each location, and multiple locations, the data points are not necessarily all random and independent.
+The measured values at each location are more likely to correlate with each other for any number of potential reasons, including measurement techniques, outside air quality, geographic distribution of vehicle types and sizes, etc.
+In short, there are many potential variables that a given location would have constant, but could affect the measured emissions.
+Therefore, each of the daily means are _not_ from the same population, so shouldn't directly be compared between locations.
+
+However, the mean emissions at each _location_ is in fact random and independent, as now the effect these previously-mentioned variables can be captured, that is, the effect on the _location_ means.
+We therefore take the mean emissions at each location:
+```{r}
+site %>%
+	`names<-`(c("Location", "Mean Emissions")) %>% 
+	kbl(digits = 1, booktabs = TRUE) %>% 
+	kable_styling(latex_options = "HOLD_position")
+```
+and can now find the mean of _these_ values to estimate the mean emission rate for all of Utah County (with 95% confidence interval):
 
 ```{r}
-#| fig.pos = "H",
-#| fig.height = 3
-
-site %>% 
-	ggplot(aes(x = DATE, y = mean_CO_g_kg)) +
-	geom_point() +
-	theme_pander() +
-	labs(
-		x = "Date",
-		y = "Mean CO Emissions (g / kg fuel)"
-	)
-```
-
-Calculating the mean of these values with a 95% confidence interval gives:
-
-```{r}
-MeanCI(site$mean_CO_g_kg) %>% 
+MeanCI(site$mean_loc) %>% 
 	bind_rows() %>% 
@@
 ```
+
+## 6.1
+
+The random, independent variable is as previously mentioned the mean CO emission rate at each _location_.
+These values can be assumed to be independent and random, but the _daily_ emission rates cannot, as the rates at each location will be correlated with each other.

### 2022-12-03 HW4/HW4.Rmd
do hw4.3

@@
 
-# 3
+# 3 (3.1)
+
+The theory behind I/M programs is good, in that vehicles with high emissions are kept from operation.
+However, as mentioned, this creates issues with equity, where these programs impact lower income individuals more than higher income individuals.
+Perhaps one of the main reasons for this is due to the age of the vehicles owned based on income.
+Newer vehicles are more fuel efficient as well as lower-emitting than older vehicles (at least as a rule), especially as emissions standards continue to get stricter.
+Newer vehicles are also significantly more expensive than older vehicles, and so are less likely to be owned by lower income individuals.
+
+But this does not wholly discount the benefit of I/M programs.
+In fact, I'd argue these programs (or at least the concept of them) are a good thing so long as equity can be addressed properly.
+Perhaps funds that would have gone to I/M programs could be used as a subsidy for low-income individuals to purchase more climate-friendly vehicles.
+Though this doesn't address the question of what to do with high-emitting vehicles directly, increasing the gas/diesel tax would discourage driving as much (lowering emissions), and then the excess funds could again be used as subsidies for low- to no-emitting vehicles.
 

### 2022-12-03 HW4/HW4.Rmd
fix hw4.3

@@
 In fact, I'd argue these programs (or at least the concept of them) are a good thing so long as equity can be addressed properly.
-Perhaps funds that would have gone to I/M programs could be used as a subsidy for low-income individuals to purchase more climate-friendly vehicles.
+Perhaps, though, funds that would have gone to I/M programs could be better used as a subsidy for low-income individuals to purchase more climate-friendly vehicles.
 Though this doesn't address the question of what to do with high-emitting vehicles directly, increasing the gas/diesel tax would discourage driving as much (lowering emissions), and then the excess funds could again be used as subsidies for low- to no-emitting vehicles.

### 2022-12-03 HW2/HW2.Rmd
finish hw2
fix hw 2.1-2.2

@@
 hdvfe <- hdv1 %>% 
-	filter(ITEM == "Average miles traveled per  gallon of fuel consumed") %>% 
-	select(YEAR, `ALL HDV`) %>% 
-	transmute(Year = YEAR, "Fuel economy (mpg)" = `ALL HDV`)
+	filter(str_detect(ITEM, "Fuel consumed") | str_detect(ITEM, "Total Rural and Urban")
+	) %>% 
+	select(YEAR, ITEM, `ALL HDV`) %>% 
+	pivot_wider(names_from = ITEM, values_from = `ALL HDV`) %>% 
+	`names<-`(c("Year", "Mmiles", "KgalFuel")) %>% 
+	transmute(Year,
+						"Fuel Economy (mpg)" = Mmiles*1000000 / (KgalFuel*1000))
 
 hdvfe %>% 
-	kbl(booktabs = TRUE, caption = "Average Fuel Economy of HDVs") %>% 
+	kbl(booktabs = TRUE, caption = "Average Fuel Economy of HDVs", digits = 1) %>% 
 	kable_styling(latex_options = "HOLD_position")
@@
 hdvem <- hdvfe %>% 
-	mutate("Fuel consumption (gpm)" = 1/`Fuel economy (mpg)`,
+	mutate("Fuel consumption (gpm)" = 1/`Fuel Economy (mpg)`,
 				 "Emissions (g CO$_2$/gal)" = co2gal["diesel"],
@@
 newpower <- oldpower - (regen/2)
+
+enginepower <- sum(tp$`Engine Power (kW)`)
 ```
@@
 This is equivalent to `r (energy <- 37.656)` kWh/gal.
-Assuming an engine efficiency of `r (eff <- 0.3)`, the total fuel needed for this drive cycle is $F = \frac{`r (oldpower) %>% round()`[\text{kWh}]/`r eff`}{`r energy %>% round(1)`[\text{kWh/gal}]} = `r (fuel1 <- oldpower/energy/eff) %>% round()`\ \text{gal}$.
+Assuming an engine efficiency of `r (eff <- 0.3)`, the total fuel needed for this drive cycle is $F = \frac{`r (enginepower) %>% round()`[\text{kWh}]/`r eff`}{`r energy %>% round(1)`[\text{kWh/gal}]} = `r (fuel1 <- enginepower/energy/eff) %>% round()`\ \text{gal}$.
 The total distance in this drive cycle is `r (totdist <- sum(tp$dist, na.rm = TRUE)) %>% round()` meters or `r (miles <- (totdist * .0006213712)) %>% round(1)` miles, giving an overall fuel economy of `r (miles/fuel1) %>% round(2)` mpg, and a fuel consumption of `r (fuel1/miles*1000) %>% round()` gallons per 1000 miles.
@@
 2022) diesel price of \$`r (price <- 4.889)`/gal, this amounts to a savings of \$`r (price*fuelsave) %>% round()` per year per truck.
+
+## 2.9
+
+There are multiple possible reasons that hybrid vehicle technology is not more widely used in drayage trucks.
+Notably, drayage trucks are often old long-haul trucks, so vehicle purchases are not often made with drayage operation in mind.
+As such, trucks are purchased to be a good fit for long-haul operation, and since long-haul trucks need a large range and are usually at high speeds, hybrid technology becomes less efficient and less practical for this purpose.
+It also takes a long time to recoup fuel costs, and trucks are usually not owned long enough for that to be a financial incentive.
 
@@
 It's possible my previous assumption that emissions would be similar between an ICE and petroleum electricity generation is misguided, in which case the numbers I calculated may change significantly.
+
+## 3.3
+Some notable sources of related emissions not included in the previous calculations are: vehicle production, transporting vehicles to dealerships or similar, refining of crude oil into gasoline, and mining operations needed for coal/natural gas.

### 2022-12-07 _targets.R
finish powerpoint

@@ -24,6 +24,7 @@ data_targets <- tar_plan(
 	utabusmiles = 15842578,
 	uvxmiles = 11,
 	uvxfleet = 21,
+	mi_kwh = 0.478,
 	
 )
 

### 2022-12-10 HW2/HW2.Rmd
update hw2 and 4 for resubmission

@@
 ---
-title: "HW 2: Energy Demand and CO2 Emissions"
+title: "HW 2: Energy Demand and CO2 Emissions RESUBMISSION"
+subtitle: "See 1.1, 2.7, 2.9"
 author: "Hayden Atchley"
@@
 ## 1.1
-The average fuel economy is
+Previously, I erroneously added the fuel economies together rather than totaling the distance and the fuel and dividing. The correct average fuel economy is
 
@@
 
+		enginepower = `Engine Power (kW)`*t/3600,
+
 		energy = power*t/3600, #kWh
@@
 
-enginepower <- sum(tp$`Engine Power (kW)`)
+enginepower <- sum(tp$enginepower, na.rm = TRUE)
 ```
@@
 ## 2.7
+Previously I used the tractive power rather than the engine power. This has been fixed.
+
 Low-sulfur diesel has an energy content of 128,488 Btu/gal according to the AFDC.
@@
 ## 2.9
+Previously I had not written a response to this question.
 

### 2022-12-10 HW4/HW4.Rmd
update hw2 and 4 for resubmission

@@
 ---
-title: "HW #4: Analyze real-world emissions data"
+title: "HW #4: Analyze real-world emissions data RESUBMISSION"
+subtitle: "See 3 & 6"
 author: Hayden Atchley
@@
 
+**Previously I didn't answer this question**
+
 The theory behind I/M programs is good, in that vehicles with high emissions are kept from operation.
@@
 
+**Previously I used _all_ samples as if they were independent rather than the means at each location.**
+
 ```{r}
@@
 	`names<-`(c("Location", "Mean Emissions")) %>% 
-	kbl(digits = 1, booktabs = TRUE) %>% 
+	kbl(digits = 1, booktabs = TRUE, linesep = "") %>% 
 	kable_styling(latex_options = "HOLD_position")

### 2022-12-13 project/Paper.Rmd
update paper outline

@@
 ---
-title: Electrification of Bus Networks in Provo, Utah (DRAFT)
+title: "Electrifying a Bus Network"
+subtitle: "A UVX Case Study"
 author: Hayden Atchley

### 2022-12-13 project/Paper.Rmd
update paper

--- 2022-11-23 project/Paper.Rmd
@@
 ---
-title: Electrification of Bus Networks in Provo, Utah (DRAFT)
+title: "Electrifying a Bus Network"
+subtitle: "A UVX Case Study"
 author: Hayden Atchley
-date: "22 November 2022"
+date: "13 December 2022"
 documentclass: article
@@
 )
+
+library(tidyverse)
+library(ggthemes)
+```
+
+```{r}
+costs <- readxl::read_xlsx("data/bus.xlsx", range = "A1:H25")
+
+fullcosts <- costs %>% 
+	mutate(
+		"50L100B" = `50lanes` + `100bus`,
+		"50L50B" = `50lanes`+ `50bus`,
+		"50LXXB" = `50lanes` + `XXbus/yr`,
+		"OPP100B" = XXopp + `100bus`,
+		"OPP50B" = XXopp + `50bus`,
+		"OPPXXB" = XXopp + `XXbus/yr`,
+		"DCFC100B" = XXdcfc + `100bus`,
+		"DCFC50B" = XXdcfc + `50bus`,
+		"DCFCXXB" = XXdcfc + `XXbus/yr`
+)
+
+scenarios <- fullcosts %>% 
+	select(-(`50lanes`:`XXbus/yr`), -which(str_detect(colnames(.), "DCFC"))) %>% 
+	pivot_longer(-Year, names_to = "Ebus", values_to = "Cost") %>% 
+	transmute(
+		Year,
+		Scenario = case_when(
+			str_detect(Ebus, "50L") ~ "Lanes",
+			str_detect(Ebus, "OPP") ~ "Chargers",
+			TRUE ~ "None"
+		),
+		Ebus = case_when(
+			str_detect(Ebus, "100B") ~ "100%",
+			str_detect(Ebus, "50B") ~ "50%",
+			str_detect(Ebus, "XXB") ~ "Gradual",
+			TRUE ~ "None"
+		),
+		CostM = Cost / 1000000
+	)
 ```
@@
 Electric motors are also efficient over a large range of rotational speeds, whereas ICEs are only efficient at a much narrower range of speeds.
+Because of these reasons, several transit agencies have deployed battery-electric (BE) buses in their fleets <><><>.
 
@@
 What truly makes this a concern is the time required to recharge an EV's battery compared to the time to refuel an ICE vehicle's tank.
-Though there are relatively fast charging standards available, as of now none of them are nearly as fast as filling up a car at a gas station [@epa-charging].
+Though there are relatively fast charging standards available, as of now none of them are nearly as fast as refueling at a gas pump [@epa-charging], especially for a vehicle as large as a bus.
 
-This charging problem, though, is not generally a concern for personal vehicles.
-Even the slowest of charging standards are able to charge at up to 1.9kW, which if charged either overnight or at a workplace for 8 hours could easily offset a daily commute of 40 miles or more [@epa-charging].
-However, for one major part of the transportation industry, namely commercial transportation, EV charging becomes a major consideration.
-Long-haul trucks, bus rapid transit (BRT), and even local bus networks may not be as well-suited for EV adoption.
-But despite the difficulty of solving the charging problem, electrification of these systems can still be a worthwhile goal, due to the many advantages of EVs.
-In fact, many transit agencies are using electric bus fleets XXXXXXXX, and others are in the process of transitioning to EVs XXXXXXX.
+Another major consideration is the source of electricity used to power EVs.
+While EVs themselves have no emissions, electricity generation often does.
+In fact, in areas such as Utah where coal is the major source of electricity <><><>, the per-mile emissions of EVs (accounting for electricity generation) can be higher than of hybrid vehicles <><><>.
+However, especially with recent legislation such as the Infrastructure Investment and Jobs Act <><> and the Inflation Reduction Act <><>, electricity generation is on the path to becoming greener overall.
 
-# Charging Solutions
+This paper, therefore, is not focused on comparing vehicles based on emissions.
+Rather, the focus is to provide a list of considerations that a transit agency should take into account when migrating to a BE bus fleet.
+Additionally, this paper presents an application of these considerations to Utah Transit Authority's (UTA's) Utah Valley Express (UVX) bus line as a case study.
 
-Developing charging infrastructure is the most significant prerequisite to deploying a fleet of EV buses.
+# Background
+
+Any transit agency operating a bus network will need to take into account several considerations.
+-fuel cost
+-fleet size
+-routing
+-fleet rotation?
+(addt'l for ev buses)
+-charging
+
+Both electric and fuel-powered vehicles have limited range, but it is generally much faster to add fuel than to recharge a battery.
+Because of this, charging considerations become paramount in any deployment of BE buses.
 There are currently several options available for EV charging.
 Level 1 and 2 charging as well as DC fast charging use a plug-in connector to transfer power [@epa-charging], and other options such as wireless inductive charging exist and are being developed upon [@Klontz1993; @Panchal2018].
-Perhaps somewhat unique to buses are <><><fancy overhead chargers><><>, where a rail on the bus extends to make contact with an overhead power supply at a bus stop.
+Perhaps somewhat unique to buses are overhead chargers, where a rail on the bus extends to make contact with an overhead power supply at a bus stop (or vice versa, as in the OppCharge standard <><><>).
 
@@
 A personal vehicle can expect about 25 miles of range per hour of level 2 charging [@epa-charging].
-A bus, which weighs several times that of a personal vehicle [@car-weights], would get a fraction of that range (since work/energy is proportional to mass).
+A bus, which weighs several times that of a personal vehicle [@car-weights], would get a fraction of that range.
 Many bus routes run for most of the day, and even with an 8-hour daily level 2 charge a bus may only gain about 50 miles of daily range.
@@
 
-DC fast charging can charge vehicles on the order of 10 times as fast as level 2 charging [@epa-charging].
+DC fast charging, on the other hand, can charge vehicles on the order of 10 times as fast as level 2 charging [@epa-charging].
 Depending on the implementation, this standard can charge as fast as 350kW.
@@
 However, standard plug-in DC fast charging overnight may not be enough on its own due to battery sizes.
-
 EV range is directly a function of battery pack size, and so a larger pack will provide more range.
 A typical electric bus with a battery pack size of 350kWh and energy consumption of about 3kWh per mile would have around 100 miles of range [@ebus-range].
-This may be sufficient for many routes much of the year, but electric heating in the winter months can drastically reduce EV range.
-Assuming a worst-case scenario of 4-6kWh per mile, this reduces the range of a 350kWh bus to around 70 miles.
-While some buses are being developed with larger battery packs [@big-ebus], these are currently not commonly in use.
+This may be sufficient for many routes, but depending on the fleet size it may not be viable to install enough charging stations to fully recharge every vehicle every day.
+
+Overhead charging allows chargers to be installed almost anywhere, so buses would potentially be able to charge mid-route.
+While this could also be accomplished with plug-in charging, the additional burden on the driver in exiting the vehicle to access the charger adds delay.
+By contrast, the OppCharge overhead charging standard is automatic and charges at up to 450kW <><><>, so 6 minutes of charging can add 10\textendash 20 miles of range to a 3kWh per mile bus.
+For a local bus route, this could enable short charging sessions at each end of the route during the day, reducing the need for overnight charging.
+
+Wireless inductive charging has the potential to obviate the need for "charging breaks" entirely.
+This technology involves installing....
+Inductive charging tends to be much slower than the other options at around 80kW <><><>, but the ability to charge while in operation can mitigate this concern.
+
+# Methods
+
+Based on these considerations....
+
+The UVX is a bus rapid transit service operated by UTA and running through Provo and Orem, Utah.
+The service uses New Flyer XDE 60 model vehicles, so to make as close of a comparison as possible, I used the New Flyer XE 60 as the electric bus model for my calculations.
+This model is nearly identical to the XDE 60, with the main difference being that the XE is BE powered and the XDE is diesel-electric <><><>.
+Table XXX lists information about the current UVX route, and Table XXX lists details of the XE 60.
+Note that the current UVX fleet size is estimated based on vehicle tracking services, and several numbers are estimated from scheduling information.
+
+TABLE
+
+Based on these values, each BE bus would need about an hour of overhead charging at 450kW per day or about 1.3 hours of 350kW plug-in charging per day.
+With only overnight (off-service) charging, a plug-in charger for every two vehicles or an overhead charger for every three vehicles would be needed.
+However, overhead charging is much easier to use during service hours.
+Traversing 11 miles at 3.4 kWh/mi requires 38 kWh, which at 450 kW would take 5.1 minutes of charging.
+If each vehicle spent this amount of time charging at each end of the route, then no off-service charging would be needed.
+Though this could theoretically be accomplished with only one charger at each end of the route, ...
+I am assuming there are two (four total).
+I am assuming as well that plug-in charging would be used mainly overnight, so 11 chargers would be required.
+
+In order to use inductive lane charging, the charge rate would need to provide 11 miles of range for every 42 minutes of charging.
+This is about 16 miles of range per hour, which at an efficiency of 3.4 kWh/mi requires a 55 kW effective charge rate.
+Assuming a relatively constant travel speed, 80 kW lane charging would need to be installed on 68% of the UVX route, or 15.1 miles, to achieve this overall rate.
+
+A cost comparison of these charging solutions is given in Table XXX.
+Many of these values are estimated based on available information.
+It is worth noting that these costs can significantly vary depending on the specific implementation, but the numbers presented represent a rough average cost.
+Note also that the expected life span of this equipment (including the vehicles) is about 10\textendash 14 years.
+
+TABLE (MO ebus and reg bus)
+
+In addition to these costs, I am estimating the fuel cost at $0.60/mi for diesel-electric vehicles and $0.10/mi for BE vehicles <><><>.
+Based on these costs, I analyzed several BE bus adoption strategies.
+The first is a baseline, with the existing fleet and no migration to BE buses.
+I also analyzed both a 100% and 50% strategy, in which BE buses instantly replace 100% and 50% of the current fleet respectively.
+Finally I analyzed a "Gradual" strategy, in which UTA replaces 2 vehicles in the current fleet with BE buses annually.
+Each of these strategies were analyzed with overhead charging, plug-in charging, and inductive lane charging.
+
+# Results
+
+An interesting result I immediately found is that given the specific scenario specifications, plug-in charging and overhead charging had nearly identical costs.
+Because of this, I averaged the cost of the two options and created a "Site Charging" scenario, as opposed to the "Inductive Lanes" scenario.
+Figures \@ref(fig:site-fig) and \@ref(fig:lanes-fig) show the cumulative cost over time of each BE bus adoption strategy.
+The shaded area represents the typical lifespan of the buses and charging equipment (10\textendash 14 years), so after this time the costs increase due to the need for equipment replacement.
+While in reality equipment would be replaced gradually rather than all at once, the cumulative effect on cost would be similar enough to not significantly affect the conclusions of this analysis.
+
+```{r}
+#| lanes-fig
+#| fig.cap = "Cumulative cost based on site charging. The shaded area indicates the typical lifespan of equipment and vehicles, so replacement costs are added at this time."
+
+scenarios %>% 
+	filter(Scenario != "Lanes") %>% 
+	ggplot(aes(x = Year, y = CostM, color = Ebus)) +
+	annotate("rect", xmin = 10, xmax = 14, ymin = -Inf, ymax = Inf,
+					 alpha = 0.2, fill = "red") +
+	geom_line(linewidth = 1) +
+	theme_bw() +
+	lims(x = c(NA,22)) +
+	labs(x = "Years of Operation",
+			 y = "Cumulative Cost (million USD)",
+			 color = "BE Bus\nStrategy",
+			 title = "Site Charging")
+```
+
+```{r}
+#| lanes-fig
+#| fig.cap = "Cumulative cost based on inductive lane charging. The shaded area indicates the typical lifespan of equipment and vehicles, so replacement costs are added at this time."
+
+scenarios %>% 
+	filter(Scenario != "Chargers") %>% 
+	ggplot(aes(x = Year, y = CostM, color = Ebus)) +
+	annotate("rect", xmin = 10, xmax = 14, ymin = -Inf, ymax = Inf,
+					 alpha = 0.2, fill = "red") +
+	geom_line(linewidth = 1) +
+	theme_bw() +
+	lims(x = c(NA,22)) +
+	labs(x = "Years of Operation",
+			 y = "Cumulative Cost (million USD)",
+			 color = "BE Bus\nStrategy",
+			 title = "Inductive Lanes")
+```
+
+Firstly, as inductive lanes are significantly more expensive than site chargers, it takes a longer time for any BE adoption strategy to save money compared to the baseline.
+In fact, no strategy is cheaper than the baseline before the estimated life span of the vehicles in the inductive lanes scenario.
+Though even the current diesel-electric buses would need to be replaced at that time, so ultimately the inductive lanes provide a reasonable solution after XXX years.
+The "Site Charging" scenario pays itself off much more quickly, especially in the 100% replacement strategy.
+
+It is important to note that UTA already has a fleet of diesel-electric buses, so capital costs are only incurred for new BE buses and charging infrastructure.
+This is why the "Gradual" strategy increases in cost so quickly; new buses are being purchased each year, but a large portion of the fleet still has a high fuel cost per mile until most of the vehicles are replaced.
+Incidentally, after 11 years, when all vehicles have been replaced in the "Gradual" strategy, the difference between the "Gradual" and "100%" strategies is purely the difference in fuel costs, which is quite substantial.
+
+- overhead/plugin about the same
+- site charging > lanes (for UVX)
+- economically viable
+
+# Limitations and Conclusions
+
+This analysis shows that the adoption of BE buses is economically viable, even when non-EV buses are already in service. (evidenced as well by agencies adopting ev buses, UTA as well)
+
+However, several significant limitations apply to this research.
+-variable costs
+-estimates
+-only UVX
+
+Conclude that even with these limitations, transit agencies should look at BE buses. Also emissions.
 

### 2022-12-13 R/data_stuff.R

@@ -1,70 +1,70 @@
-library(tidyverse)
-library(ggthemes)
-
-costs <- readxl::read_xlsx("data/bus.xlsx", range = "A1:H25")
-
-fullcosts <- costs %>% 
-	mutate(
-		"50L100B" = `50lanes` + `100bus`,
-		"50L50B" = `50lanes`+ `50bus`,
-		"50LXXB" = `50lanes` + `XXbus/yr`,
-		"OPP100B" = XXopp + `100bus`,
-		"OPP50B" = XXopp + `50bus`,
-		"OPPXXB" = XXopp + `XXbus/yr`,
-		"DCFC100B" = XXdcfc + `100bus`,
-		"DCFC50B" = XXdcfc + `50bus`,
-		"DCFCXXB" = XXdcfc + `XXbus/yr`
-)
-
-scenarios <- fullcosts %>% 
-	select(-(`50lanes`:`XXbus/yr`), -which(str_detect(colnames(.), "DCFC"))) %>% 
-	pivot_longer(-Year, names_to = "Ebus", values_to = "Cost") %>% 
-	mutate(
-		Scenario = case_when(
-			str_detect(Ebus, "50L") ~ "Lanes",
-			str_detect(Ebus, "OPP") ~ "Chargers",
-			TRUE ~ "None"
-		),
-		Ebus = case_when(
-			str_detect(Ebus, "100B") ~ "100%",
-			str_detect(Ebus, "50B") ~ "50%",
-			str_detect(Ebus, "XXB") ~ "Gradual",
-			TRUE ~ "None"
-		)
-	)
-
-scenarios %>% 
-	filter(Scenario != "Chargers") %>% 
-	ggplot(aes(x = Year, y = Cost, color = Ebus)) +
-	annotate("rect", xmin = 10, xmax = 14, ymin = -Inf, ymax = Inf,
-					 alpha = 0.2, fill = "red") +
-	geom_line(linewidth = 1) +
-	theme_pander() +
-	lims(x = c(NA,20)) +
-	labs(x = "Years of Operation",
-			 y = "Cumulative Cost (USD)",
-			 color = "BE Bus\nStrategy",
-			 title = "Inductive Lanes")
-
-ggsave("project/image/inductive_lanes.png",
-			 width = 7,
-			 height = 6,
-			 units = "in")
-
-scenarios %>% 
-	filter(Scenario != "Lanes") %>% 
-	ggplot(aes(x = Year, y = Cost, color = Ebus)) +
[…]

### 2022-12-14 project/Paper.Rmd
update paper

@@
 library(ggthemes)
+library(kableExtra)
 ```
@@
 This model is nearly identical to the XDE 60, with the main difference being that the XE is BE powered and the XDE is diesel-electric <><><>.
-Table XXX lists information about the current UVX route, and Table XXX lists details of the XE 60.
+Table \@ref(tab:uvx) lists information about the current UVX route, and Table \@ref(tab:xe) lists details of the XE 60.
 Note that the current UVX fleet size is estimated based on vehicle tracking services, and several numbers are estimated from scheduling information.
 
-TABLE
+```{r}
+#| uvx
+
+uvx <- read_csv("data/uvx.csv", col_names = FALSE)
+
+uvx %>% 
+	kbl(
+		align = 'lrl',
+		col.names = c("Utah Valley Express (830X)", "", ""),
+		booktabs = TRUE,
+		caption = "UVX Route Information") %>% 
+	kable_styling(latex_options = "hold_position")
+```
+
+```{r}
+#| xe
+
+xe <- read_csv("data/xe60.csv", col_names = FALSE)
+
+xe %>% 
+	kbl(
+		align = 'lrl',
+		col.names = c("New Flyer XE 60", "", ""),
+		booktabs = TRUE,
+		caption = "New Flyer XE 60 Information") %>% 
+	kable_styling(latex_options = "hold_position")
+```
 
@@
 
-A cost comparison of these charging solutions is given in Table XXX.
+A cost comparison of these charging solutions is given in Table \@ref(tab:costs).
 Many of these values are estimated based on available information.
@@
 
-TABLE (MO ebus and reg bus)
+```{r}
+#| costs
+
+tribble(
+	~item, ~cap, ~mo,
+	"XE 60 Bus", 750000, 7000,
+	"XDE 60 Bus", 1000000, 50000,
+	"Plug-in Charger", 20000, 400,
+	"Overhead Charger", 40000, 800,
+	"Inductive Lanes (per mile)", 322000, 6400
+) %>% 
+	kbl(
+		col.names = c("", "Capital Cost (USD)", "Maintenance Cost (USD/yr)"),
+		caption = "Fleet Costs",
+		booktabs = TRUE
+	) %>% 
+	kable_styling(latex_options = "hold_position")
+```
 
@@
 Because of this, I averaged the cost of the two options and created a "Site Charging" scenario, as opposed to the "Inductive Lanes" scenario.
-Figures \@ref(fig:site-fig) and \@ref(fig:lanes-fig) show the cumulative cost over time of each BE bus adoption strategy.
+Figures \@ref(fig:site) and \@ref(fig:lanes) show the cumulative cost over time of each BE bus adoption strategy.
 The shaded area represents the typical lifespan of the buses and charging equipment (10\textendash 14 years), so after this time the costs increase due to the need for equipment replacement.
@@
 ```{r}
-#| lanes-fig
-#| fig.cap = "Cumulative cost based on site charging. The shaded area indicates the typical lifespan of equipment and vehicles, so replacement costs are added at this time."
+#| site,
+#| fig.height = 4,
+#| fig.cap = "Cumulative cost based on site charging. The shaded area indicates the typical lifespan of equipment and vehicles, so replacement costs are added at this time.",
+#| fig.pos = "h"
 
@@
 ```{r}
-#| lanes-fig
-#| fig.cap = "Cumulative cost based on inductive lane charging. The shaded area indicates the typical lifespan of equipment and vehicles, so replacement costs are added at this time."
+#| lanes,
+#| fig.height = 4,
+#| fig.cap = "Cumulative cost based on inductive lane charging. The shaded area indicates the typical lifespan of equipment and vehicles, so replacement costs are added at this time.",
+#| fig.pos = "h"
 
@@
 In fact, no strategy is cheaper than the baseline before the estimated life span of the vehicles in the inductive lanes scenario.
-Though even the current diesel-electric buses would need to be replaced at that time, so ultimately the inductive lanes provide a reasonable solution after XXX years.
-The "Site Charging" scenario pays itself off much more quickly, especially in the 100% replacement strategy.
+Though even the current diesel-electric buses would need to be replaced at that time, so ultimately the inductive lanes provide a reasonable solution after 15\textendash 20 years.
+The "Site Charging" scenario pays itself off much more quickly (11\textendash 13 years), especially in the 100% replacement strategy.
 
 It is important to note that UTA already has a fleet of diesel-electric buses, so capital costs are only incurred for new BE buses and charging infrastructure.
-This is why the "Gradual" strategy increases in cost so quickly; new buses are being purchased each year, but a large portion of the fleet still has a high fuel cost per mile until most of the vehicles are replaced.
-Incidentally, after 11 years, when all vehicles have been replaced in the "Gradual" strategy, the difference between the "Gradual" and "100%" strategies is purely the difference in fuel costs, which is quite substantial.
-
-- overhead/plugin about the same
-- site charging > lanes (for UVX)
-- economically viable
+This is why the "Gradual" strategy increases in cost so quickly; new buses are being purchased each year, but a large portion of the fleet still has a high fuel cost per mile and maintenance cost until most of the vehicles are replaced.
 
@@
 Conclude that even with these limitations, transit agencies should look at BE buses. Also emissions.
+- site charging > lanes (for UVX)
+- economically viable
+
+\newpage
 

### 2022-12-14 project/Paper.Rmd
finish paper

@@
 Electric motors are also efficient over a large range of rotational speeds, whereas ICEs are only efficient at a much narrower range of speeds.
-Because of these reasons, several transit agencies have deployed battery-electric (BE) buses in their fleets <><><>.
+Because of these reasons, several transit agencies have deployed battery-electric (BE) buses in their fleets [@Li2016; @Topal2018; @Zhang2022].
 
@@
 While EVs themselves have no emissions, electricity generation often does.
-In fact, in areas such as Utah where coal is the major source of electricity <><><>, the per-mile emissions of EVs (accounting for electricity generation) can be higher than of hybrid vehicles <><><>.
-However, especially with recent legislation such as the Infrastructure Investment and Jobs Act <><> and the Inflation Reduction Act <><>, electricity generation is on the path to becoming greener overall.
+In fact, in areas such as Utah where coal is the major source of electricity, the per-mile emissions of EVs (accounting for electricity generation) can be higher than of hybrid vehicles.
+However, especially with recent legislation such as the @iija and the @ira, electricity generation is on the path to becoming greener overall.
 
@@
 Any transit agency operating a bus network will need to take into account several considerations.
--fuel cost
--fleet size
--routing
--fleet rotation?
-(addt'l for ev buses)
--charging
+These include fleet size, routing, fuel costs, and potentially a rotation of vehicles for maintenance.
+These considerations apply regardless of the type of vehicle used, but BE buses bring additional concerns for vehicle range and charging capacity.
 
@@
 Level 1 and 2 charging as well as DC fast charging use a plug-in connector to transfer power [@epa-charging], and other options such as wireless inductive charging exist and are being developed upon [@Klontz1993; @Panchal2018].
-Perhaps somewhat unique to buses are overhead chargers, where a rail on the bus extends to make contact with an overhead power supply at a bus stop (or vice versa, as in the OppCharge standard <><><>).
+Perhaps somewhat unique to buses are overhead chargers, where a rail on the bus extends to make contact with an overhead power supply at a bus stop (or vice versa, as in the OppCharge standard [@oppcharge]).
 
@@
 While this could also be accomplished with plug-in charging, the additional burden on the driver in exiting the vehicle to access the charger adds delay.
-By contrast, the OppCharge overhead charging standard is automatic and charges at up to 450kW <><><>, so 6 minutes of charging can add 10\textendash 20 miles of range to a 3kWh per mile bus.
+By contrast, the OppCharge overhead charging standard is automatic and charges at up to 450kW [@oppcharge], so 6 minutes of charging can add 10\textendash 20 miles of range to a 3kWh per mile bus.
 For a local bus route, this could enable short charging sessions at each end of the route during the day, reducing the need for overnight charging.
@@
 Wireless inductive charging has the potential to obviate the need for "charging breaks" entirely.
-This technology involves installing....
-Inductive charging tends to be much slower than the other options at around 80kW <><><>, but the ability to charge while in operation can mitigate this concern.
+This technology involves installing charging infrastructure in the road itself, to wirelessly charge vehicles that drive over it.
+Inductive charging tends to be much slower than the other options at around 80kW [@Chen2018], but the ability to charge while in operation, as opposed to stationary charging, can mitigate this concern.
 
@@
 
-Based on these considerations....
-
-The UVX is a bus rapid transit service operated by UTA and running through Provo and Orem, Utah.
+In order to assess the viability of these charging solutions, I am applying these considerations to UTA's UVX bus route.
+The UVX is a bus rapid transit (BRT) service running through Provo and Orem, Utah.
 The service uses New Flyer XDE 60 model vehicles, so to make as close of a comparison as possible, I used the New Flyer XE 60 as the electric bus model for my calculations.
-This model is nearly identical to the XDE 60, with the main difference being that the XE is BE powered and the XDE is diesel-electric <><><>.
-Table \@ref(tab:uvx) lists information about the current UVX route, and Table \@ref(tab:xe) lists details of the XE 60.
+This model is nearly identical to the XDE 60, with the main difference being that the XE is BE powered and the XDE is diesel-electric [@newflyer].
+Table \@ref(tab:uvx) lists information about the current UVX route, and Table \@ref(tab:xe) lists details of the XE 60 [@newflyerxe].
 Note that the current UVX fleet size is estimated based on vehicle tracking services, and several numbers are estimated from scheduling information.
@@
 uvx %>% 
+	mutate(across(.fns = as.character)) %>% 
 	kbl(
@@
 		booktabs = TRUE,
-		caption = "UVX Route Information") %>% 
+		caption = "UVX Route Information",
+		linesep = "") %>% 
 	kable_styling(latex_options = "hold_position")
@@
 xe %>% 
+	mutate(across(.fns = as.character)) %>% 
 	kbl(
@@
 If each vehicle spent this amount of time charging at each end of the route, then no off-service charging would be needed.
-Though this could theoretically be accomplished with only one charger at each end of the route, ...
-I am assuming there are two (four total).
-I am assuming as well that plug-in charging would be used mainly overnight, so 11 chargers would be required.
+Though this could theoretically be accomplished with only one charger at each end of the route, it would be more efficient (and probably more likely) for this to be doubled with two chargers at each end, when multiple buses are waiting to charge.
+This would also for redundancy in case of a charger malfunction.
+Plug-in charging, however, is more likely to be used mainly outside service hours, so the full fleet would need to be charged each night.
+This would require 11 chargers for 21 vehicles.
 
@@
 A cost comparison of these charging solutions is given in Table \@ref(tab:costs).
-Many of these values are estimated based on available information.
+Many of these values are estimated based on available information [@Topal2018; @DiNello2021; @Wortrich2022].
 It is worth noting that these costs can significantly vary depending on the specific implementation, but the numbers presented represent a rough average cost.
@@
 		caption = "Fleet Costs",
-		booktabs = TRUE
+		booktabs = TRUE,
+		align = 'lcc'
 	) %>% 
@@
 
-In addition to these costs, I am estimating the fuel cost at $0.60/mi for diesel-electric vehicles and $0.10/mi for BE vehicles <><><>.
+In addition to these costs, I am estimating the fuel cost at \$0.60/mi for diesel-electric vehicles and \$0.10/mi for BE vehicles [@Topal2018].
 Based on these costs, I analyzed several BE bus adoption strategies.
@@
 #| site,
-#| fig.height = 4,
+#| fig.height = 3.5,
 #| fig.cap = "Cumulative cost based on site charging. The shaded area indicates the typical lifespan of equipment and vehicles, so replacement costs are added at this time.",
-#| fig.pos = "h"
+#| fig.pos = "H"
 
@@
 #| lanes,
-#| fig.height = 4,
+#| fig.height = 3.5,
 #| fig.cap = "Cumulative cost based on inductive lane charging. The shaded area indicates the typical lifespan of equipment and vehicles, so replacement costs are added at this time.",
-#| fig.pos = "h"
+#| fig.pos = "H"
 
@@
 Though even the current diesel-electric buses would need to be replaced at that time, so ultimately the inductive lanes provide a reasonable solution after 15\textendash 20 years.
-The "Site Charging" scenario pays itself off much more quickly (11\textendash 13 years), especially in the 100% replacement strategy.
+The "Site Charging" scenario pays itself off much more quickly, especially in the 100% replacement strategy (11\textendash 13 years).
 
@@
 
-This analysis shows that the adoption of BE buses is economically viable, even when non-EV buses are already in service. (evidenced as well by agencies adopting ev buses, UTA as well)
+This analysis shows that the adoption of BE buses is economically viable, even when non-EV buses are already in service.
+Certainly as current vehicles reach the end of their life there is a very strong case to replace them with BE buses, as fuel and maintenance costs are lower, even accounting for the charging infrastructure.
+However, several significant limitations apply to this research.
+The costs used in the analysis are at best rough approximations of actual real-world costs, so any agency looking to adopt BE buses should conduct a more thorough cost analysis than the one presented here.
+Additionally, this analysis looked specifically at the UVX service, which is not necessarily representative of local bus routes, or even other BRT services.
 
-However, several significant limitations apply to this research.
--variable costs
--estimates
--only UVX
-
-Conclude that even with these limitations, transit agencies should look at BE buses. Also emissions.
-- site charging > lanes (for UVX)
-- economically viable
+Even with these limitations, though, BE buses present a compelling solution for transit agencies.
+This is evidenced by their growing adoption [@Li2016; @Topal2018; @Zhang2022], including in UTA's own fleet [@uta2021review].
+In adopting BE fleets and building charging infrastructure, this paper generally recommends site charging solutions (plug-in and/or overhead) to inductive lane charging based on cost, but again, a more thorough analysis should be performed in each specific circumstance.
 
